---
title: "Classification of palmetto species by their morphological characteristics"
subtitle: "ESM 244 - Advanced Daya Analysis : Homework 2, task 2"
author: "Carmen Galaz-García"
date: "1/29/2021"
output: 
  html_document:
    toc: true  # make a table of contents
    toc_float: true  # table of contents as sidebar
    number_sections: true  
    code_folding: hide  # makes hide/show for code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning=FALSE)

library(tidyverse)
library(broom)
library(kableExtra)  # create nice tables

```

# Introduction 

In this report we use binary logistic regression to test the feasibility of using morphological characteristics to classify whether a palmetto plant is species *serenoa repens* or species *sabal etonia*. 
The variables we use for this classification are the height (cm), widest length of the canopy (cm), widest width of the canopy perpendicular to the canopy length (cm) and number of green leaves (n) in a palmetto plant.
Data was collected by the Archbold Biological Station in south-central Florida from 1981 through 1997 then again in 2001 and 2017.
 
 
All analyses are in R version 4.0.2 using RStudio Version 1.3.1093.

# Data exploration
```{r}
# --- DATA SELECTION ---- 

raw_palmetto <- read_csv("palmetto.csv")

palmetto <- raw_palmetto %>% 
  select(species, height:green_lvs) %>% 
  mutate(species = case_when( species==1 ~ "serenoa repens",   # update species names
                              species==2 ~ "sabal etonia"))

palmetto$species<- as.factor(palmetto$species) # convert to factor
#levels(palmetto$species)    # check factor levels
```


```{r}
# --- GRAPH 1: CANOPY DIMENSIONS ---

ggplot(data=palmetto, aes(x=length, y=width))+
  geom_point( cex=0.3, alpha=0.2)+    # point size and transparency
  facet_wrap(~species)+     # divide by species
  labs( x= "length (cm)",   # add labels and title
        y= "width (cm)",
       title = expression(paste(
       "Canopy dimensions in ",
       italic("sabal etonia"),
       " and ",
       italic("serenoa repens"),
       " individuals"))
        )+
  theme_light()
  
```

**Figure 1.** Relationship between widest length of the canopy (cm) and widest width of the canopy perpendicular to the canopy length (cm) of the palmetto plant species *sabal etonia* and *serenoa repens*. Each point indicates measurements of an individual plant. Data: Abrahamson (2019)


```{r}
# --- GRAPH 2: NUMBER OF LEAVES HISTOGRAM  ---

ggplot(data=palmetto, aes(x=green_lvs))+
  geom_histogram(binwidth = 1)+  # update number of bins
  facet_wrap(~species)+                     # divide by species
  labs( x = "number of green leaves (n)",   # add labels and title
        y = "count",
        title = expression(paste(
       "Number of green leaves in ",
       italic("sabal etonia"),
       " and ",
       italic("serenoa repens"),
       " individuals"))
       )+
  theme_light()
```

**Figure 2.** Histograms of number of green leaves (n) in the palmetto plant species *sabal etonia* and *serenoa repens*. Data: Abrahamson (2019)


```{r}
# --- GRAPH 3: 
summary_palmetto <- palmetto %>% 
  group_by(species) %>% 
  summarize(mean = mean(height, na.rm=TRUE),
            sd = sd(height, na.rm=TRUE),
            #median = median(weight, na.rm=TRUE),
            n = n()
            )

ggplot(data=palmetto, aes(x=species, y=height))+
  geom_boxplot()+
  geom_jitter(color="forestgreen", size=0.4, alpha=0.2) +
  labs(y = "height (cm)",
        title = expression(paste(
       "Height of ",
       italic("sabal etonia"),
       " and ",
       italic("serenoa repens"),
       " individuals"))
       )+
  theme_minimal()

```

**Figure 3.** Height (cm) of the palmetto plant species *sabal etonia* and *serenoa repens*. Each green point indicates the height of an individual plant. 
Box endpoints indicate the 25th and 75th percentile values, the black line within the box indicate the median for each species. 
Black points at the end of the boxes indicate possible outliers. 
Data: Abrahamson (2019)


# Binary logistic regression

```{r}
# ---- BINARY LOGISTIC REGRESSION ---- 

# -- create model
blr_palmetto <- glm(species ~ height + length + width + green_lvs,
                            data = palmetto,
                            family = "binomial") #specify model is BLR

# -- table with model information 
blr_model <- tidy(blr_palmetto) %>% 
  mutate(pvalue = "< 0.001") %>%     # after looking at p-values, they are all <0.001
  mutate(term = c("y-intercept",     # update variable names
                  "height (cm)", 
                  "length (cm)", 
                  "width (cm)", 
                  "number of green leaves (n)")
         ) %>% 
  select(!statistic & !p.value)%>%   # exclude statistic and p-value columns
  kable( col.names = c("Variable",   # turn into kable and update column names
                        "Coefficient", 
                        "Standard error",
                        "p-value") 
                         ) %>% 
   kable_styling(full_width = FALSE) 

blr_model
```

## Model accuracy

```{r}
# ---- MODEL ACCURACY ---- 

results_blr <- blr_palmetto %>% 
  broom::augment(type.predict="response") %>% # calculate probability for each outcome
  select(species,.fitted) %>%                 # select species and probability columns only
  # check if classifcation is correct
  mutate(classification = case_when( .fitted>0.5 & species=="serenoa repens"~ "True",   
                                     .fitted<=0.5 & species=="serenoa repens"~ "False",
                                     .fitted>0.5 & species=="sabal etonia"~ "False",
                                     .fitted<=0.5 & species=="sabal etonia"~ "True")) %>% 
  count(species, classification) %>%    
  mutate(percentage = case_when(     # add column with percentages 
    species == "sabal etonia" ~ scales::percent(n/(454+5701),accuracy = 0.01),
    species == "serenoa repens" ~ scales::percent(n/(564+5548),accuracy = 0.01))
    ) %>% 
  kable( col.names = c("Species",    # create kable table, update column names
                        "Correctly classified by model", 
                        "Count",
                       "Percentage") 
                         ) %>% 
   kable_styling(full_width = FALSE)

results_blr
```

# Citation
Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative.
https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5